In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-30 paper/main.tex
Updates from Overleaf

\title{In-silico molecular enrichment and clearance of the human intracranial space}
\author[1]{Marius Causemann}
\author[1]{Miroslav Kuchta}
\author[2]{Rami Masri}
\author[1,3,*]{Marie E. Rognes }
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[2]{Brown University, Providence, Rhode Island, USA}
\affil[3]{K. G. Jebsen Centre for Brain Fluid Research, University of Oslo, Norway}
\affil[*]{meg@simula.no}

@@

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to
  human brain health, with alterations often linked to
  disease and functional impairment, and with distinct opportunities
  for personalized diagnostics and treatment. However, our
  understanding of these mechanisms and their interplay remains
  incomplete, in part due to the complexity of integrating insights
  across scales, between species and from different modalities. Here,
  we combine mixed-dimensional modelling, multi-modal magnetic 
  resonance images, and high performance computing to construct and 
  explore a high-fidelity in-silico model of human intracranial molecular enrichment. 
  This model predicts the temporo-spatial spreading of a solute
  within an image-derived geometric representation of the
  subarachnoid space, ventricular system and brain parenchyma,
  including networks of surface perivascular spaces
  (PVSs). Our findings highlight the significant impact of
  cerebrospinal fluid (CSF) production and intracranial pulsatility on
  molecular enrichment following intrathecal tracer injection. We
  demonstrate that low-frequency vasomotion induces moderate CSF flow
  in surface PVS networks that substantially enhances tracer
  enrichment, and that impaired enrichment is a direct
  consequence of enlarged PVSs. This openly available technology
  platform thus provides an opportunity for integrating separate
  observations on diffusion in neuropil, vascular dynamics,
  intracranial pulsatility, CSF production, and efflux, and for
  exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

% Main findings and outlook
By exploring this high-dimensional parameter space, we propose that
the balance between CSF production and intracranial pulsatility is key
to shaping the large-scale features of intracranial enrichment
patterns, with the potential to span a wide range of individual and
cohort variability. Moreover, we predict that CSF production, cardiac-
and respiratory pulsatility is not sufficient to explain early
perivascular enrichment, but that fluid flow induced by low-frequency
vasomotion in surface periarterial spaces (on the order of 13 \textmu
m/s) is sufficient, even in the absence of a structural
compartmentalization of the PVS. Conversely, enlarged PVSs in the SAS
will cause a substantial reduction in cardiac- and vasomotion-driven
flow velocities, strongly delay perivascular transport, and thus
impair intracranial enrichment. These findings transfer, reconcile,
integrate and extend insights from clinical, experimental, and
theoretical studies, and lay a new foundation for in-silico studies of
personalized intrathecal drug delivery and brain clearance.

@@

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure1.png}
     \caption{A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}
(\Cref{fig:results1}D). Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}, we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}E--G). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated several centimeters into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}I). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}H).

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs. A key observation is that the balance between CSF production and
intracranial pulsatility significantly shapes the large-scale features
of molecular enrichment, with e.g.~ventricular tracer reflux after
intrathecal injection in the absence of CSF production, and preference
towards anterior brain regions with reduced pulsatility. Indeed, the
simulated distribution patterns cover a substantial range of the
individual variations observed clinically \cite{ringstad2018brain}. In
terms of perivascular pathways, we find that even moderate CSF flow on
the order of 0.1--1.0 \textmu m/s in the surface PVSs results in
earlier tracer enrichment around major cerebral arteries. This effect
is more pronounced -- with up to a three-hour difference in time of
arrival -- with PVS flow rates of approximately 10 \textmu m/s. Our
models of CSF flow in perivascular networks, based on first principles
and asymptotic analysis, predict net flow of such magnitudes and both
antegrade and retrograde flow within the perivascular network,
occurring with or without pressure differences at the network
ends. The PVS may thus function as highways facilitating rapid
transport, even in the absence of structural barriers, though a sharp
concentration difference between the PVS and surrounding tissues is
unlikely without such barriers. Finally, dilated PVS will cause a
substantial reduction in cardiac- and vasomotion-driven flow
velocities, which in turn impedes perivascular transport.

@@

% Tracer enrichment predictions
While molecular enrichment of the brain and surrounding CSF spaces
have been extensively studied in animal models, especially in murine
models in connection with studies of the glymphatic system, there are
fewer reports of contrast enrichment in human subjects over 0--24h. We
therefore mainly compare our in-silico predictions with the series of
papers by Eide and colleagues~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2021sleep, eide2024functional} and Watts et
al~\cite{watts2019measuring}. Ringstad et al.~\cite{ringstad2018brain}
report of tracer enrichment in a centripetal pattern, primarily in
regions near large cerebral surface arteries, peaking in the CSF
spaces between 6 and 9 hours, with parenchymal tracer content still
increasing after 24 hours, but with large individual
differences. Watts et al~\cite{watts2019measuring} report of a similar
enrichment pattern and a concentration peak in the SAS of $\approx$0.5
mmol/l (relative to a total amount of 1.0 mmol) occuring after 10 to 15 hours. Our in-silico enrichment
patterns agree with these observations, though with an earlier rather than later peak in the CSF spaces (2 hours after first tracer appearance at the craniocervical junction). In particular, we note that the substantial individual variation observed clinically is comparable to the variation in in-silico enrichment patterns associated with reduced CSF production or reduced intracranial pulsatility. We find that about 20\% of the tracer reaching the cranium enters the parenchyma, which aligns with the reported peak concentration values of $0.5\,$mmol/l in the SAS and $0.1\,$mmol/l in the parenchyma reported by Watts et al~\cite{watts2019measuring}.
In terms of perivascular tracer
enrichment, Eide and Ringstad~\cite{eide2024functional} report of
first-time tracer appearance in the PVS of the MCA after 37.8 $\pm$
47.0 (M1), 53.1 $\pm$ 50.5 (M2) and 82.6 $\pm$ 61.5 (M3) min after injection, thus with just a 15.3 min delay between M2 and M1. Comparing with our FTAs in the MCA segments, we note that the baseline model predicts a more-than 3 hour difference in FTA between the M1 and M2. Instead, the high PVS flow scenario predicts a 24 min delay between these segments.
Moreover, the high PVS flow scenario predicts FTAs of
84--108 min (M1) and 108--132 min (M2) after tracer appearance at the craniocervical junction, which overlaps with the previous ranges if one assumes a tracer travel time from the point of injection to the craniocervical junction of $13\,$min (as reported by Eide and Ringstad \cite{eide2024functional}). We interpret these results as indicative of a net CSF flow in human surface PVS beyond what would be induced by CSF production and cardiac peristalsis alone.
% M2 - M1 (mean) ~ 15.3 min

@@

$\Lambda_a$, linearly decreasing over the corresponding cell, to account for the transition from explicitly resolved

>>>>

\title{In-silico molecular enrichment and clearance of the human intracranial space}
\author[1]{Marius Causemann}
\author[1]{Miroslav Kuchta}
\author[2]{Rami Masri}
\author[1,3,*]{Marie E. Rognes }
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[2]{Division of Applied Mathematics, Brown University, Providence, Rhode Island, USA}
\affil[3]{K. G. Jebsen Centre for Brain Fluid Research, University of Oslo, Norway}
\affil[*]{meg@simula.no}

@@

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to
  human brain health, with alterations often linked to
  disease and functional impairment, and with distinct opportunities
  for personalized diagnostics and treatment. However, our
  understanding of these mechanisms and their interplay remains
  incomplete, in part due to the complexity of integrating insights
  across scales, between species and from different modalities. Here,
  we combine mixed-dimensional modelling, multi-modal magnetic 
  resonance images, and high performance computing to construct and 
  explore a high-fidelity in-silico model of human intracranial molecular enrichment. 
  This model predicts the temporo-spatial spreading of a solute
  within an image-derived geometric representation of the
  subarachnoid space, ventricular system and brain parenchyma,
  including networks of surface perivascular spaces
  (PVSs). Our findings highlight the significant impact of
  cerebrospinal fluid (CSF) production and intracranial pulsatility on
  molecular enrichment following intrathecal tracer injection. We
  demonstrate that low-frequency vasomotion induces moderate CSF flow
  in surface PVS networks which substantially enhances tracer
  enrichment, and that impaired enrichment is a direct natural
  consequence of enlarged PVSs. This openly available technology
  platform thus provides an opportunity for integrating separate
  observations on diffusion in neuropil, vascular dynamics,
  intracranial pulsatility, CSF production, and efflux, and for
  exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

% Main findings and outlook
By exploring this high-dimensional parameter space, we propose that
the balance between CSF production and intracranial pulsatility is key
to shaping the large-scale features of intracranial enrichment
patterns, with the potential to span a wide range of individual and
cohort variability. Moreover, we predict that CSF production, cardiac-
and respiratory pulsatility is not sufficient to explain early
perivascular enrichment, but that fluid flow induced by low-frequency
vasomotion in surface periarterial spaces (on the order of 10 \textmu
m/s) is sufficient, even in the absence of a structural
compartmentalization of the PVS. Conversely, enlarged PVSs in the SAS
will cause a substantial reduction in cardiac- and vasomotion-driven
flow velocities, strongly delay perivascular transport, and thus
impair intracranial enrichment. These findings transfer, reconcile,
integrate and extend insights from clinical, experimental, and
theoretical studies, and lay a new foundation for in-silico studies of
personalized intrathecal drug delivery and brain clearance.

@@

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure1.png}
     \caption{A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}(\Cref{fig:results1}D), we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}E--G). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated substantially into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}I). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}H).

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs. A key observation is that the balance between CSF production and
intracranial pulsatility significantly shapes the large-scale features
of molecular enrichment, with e.g.~ventricular tracer reflux after
intrathecal injection in the absence of CSF production, and preference
towards anterior brain regions with reduced pulsatility. Indeed, the
simulated distribution patterns cover a substantial range of the
individual variations observed clinically \cite{ringstad2018brain}. In
terms of perivascular pathways, we find that even moderate CSF flow on
the order of 0.1--1.0 \textmu m/s in the surface PVSs results in
earlier tracer enrichment around major cerebral arteries. This effect
is more pronounced -- with up to a three-hour difference in time of
arrival -- with mean PVS flow rates of approximately 10 \textmu m/s. Our
models of CSF flow in perivascular networks, based on first principles
and asymptotic analysis, predict net flow of such magnitudes and both
antegrade and retrograde flow within the perivascular network,
occurring with or without pressure differences at the network
ends. The PVS may thus function as highways facilitating rapid
transport, even in the absence of structural barriers, though a sharp
concentration difference between the PVS and surrounding tissues is
unlikely without such barriers. Finally, dilated PVS will cause a
substantial reduction in cardiac- and vasomotion-driven flow
velocities, which in turn impedes perivascular transport.

@@

% Tracer enrichment predictions
While molecular enrichment of the brain and surrounding CSF spaces
have been extensively studied in animal models, especially in murine
models in connection with studies of the glymphatic system, there are
fewer reports of contrast enrichment in human subjects over 0--24h. We
therefore mainly compare our in-silico predictions with the series of
papers by Eide and colleagues~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2021sleep, eide2024functional} and Watts et
al~\cite{watts2019measuring}. Ringstad et al.~\cite{ringstad2018brain}
report of tracer enrichment in a centripetal pattern, primarily in
regions near large cerebral surface arteries, peaking in the CSF
spaces between 6 and 9 hours, with parenchymal tracer content still
increasing after 24 hours, but with large individual
differences. Watts et al~\cite{watts2019measuring} report of a similar
enrichment pattern and a concentration peak in the SAS of $\approx$0.5
mmol/l (relative to a total amount of 1.0 mmol) occuring after 10 to 15 hours. Our in-silico enrichment
patterns agree with these observations, though with an earlier rather than later peak in the CSF spaces (2 hours after first tracer appearance at the craniocervical junction). In particular, we note that the substantial individual variation observed clinically is comparable to the variation in in-silico enrichment patterns associated with reduced CSF production or reduced intracranial pulsatility. We find that about 20\% of the tracer reaching the cranium enters the parenchyma, which aligns with the reported peak concentration values of $0.5\,$mmol/l in the SAS and $0.1\,$mmol/l in the parenchyma reported by Watts et al~\cite{watts2019measuring}.
In terms of perivascular tracer
enrichment, Eide and Ringstad~\cite{eide2024functional} report of
first-time tracer appearance in the PVS of the MCA after 37.8 $\pm$
47.0 (M1), 53.1 $\pm$ 50.5 (M2) and 82.6 $\pm$ 61.5 (M3) min after injection, thus with just a 15.3 min delay between M2 and M1. Comparing with our FTAs in the MCA segments, we note that the baseline model predicts a more-than 3 hour difference in FTA between the M1 and M2. In contrast, the high PVS flow scenario predicts a 24 min delay between these segments.
Moreover, the high PVS flow scenario predicts FTAs of
84--108 min (M1) and 108--132 min (M2) after tracer appearance at the craniocervical junction.
Assuming a tracer travel time from the point of injection to the craniocervical junction of $13\,$min (as reported by Eide and Ringstad \cite{eide2024functional}), our predictions fall at the upper end of the clinical ranges. This can partly be attributed to the limited caudal extend of our PVS network, with the lowest point located approximately $4\,$cm above the craniocervical junction due to imaging constrains.
We interpret these results as indicative of a net CSF flow in human surface PVS beyond what would be induced by CSF production and cardiac peristalsis alone.
% M2 - M1 (mean) ~ 15.3 min

@@

$\Lambda_a$, linearly decreasing to the baseline permeability at the next network node ($\approx0.4\,$mm), to account for the transition from explicitly resolved
